---
title: "TB: exploring the case of TB elimination with a simple compartmental model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ending Tuberculosis (TB) and elimination are terms now more present than ever in the Global Health agenda. Is it achievable? how and when could this happen?. All this is yet uncertain, and an ongoing discussion in the field. In this practical, we intend to illustrate important concepts of TB epidemiology by setting up a simple compartmental model and trying to assess what would be required to End TB goals as set by WHO and/or elimination thresholds.

##Core concepts The following concepts will be developed further during the practical:

-   Compartmental Models
-   Flow diagrams
-   Long-term infection dynamics
-   Natural History of TB
-   TB control strategies

##Required packages

```{r}
#install.packages("deSolve", dep=TRUE)
#install.packages("gridExtra", dep = TRUE)
#install.packages("ggplot2", dep = TRUE)
```

## Part I: Set up a basic TB model

### Natural History of TB

Tuberculosis is an infectious disease caused by *Mycobacterium* *tuberculosis*. Pulmonary tuberculosis (PTB) is the most prevalent manifestation of the disease and also the form more likely to be infectious. TB is transmitted from person to person via micro droplets sprayed by an infectious individual when they cough or sneeze. Studies of household contacts and phylogenetic analysis have estimated the transmission potential of the average infectious TB case to be between 4 and 18 secondary cases per year (1).

An important feature in the timeline of TB infection is the latent period. The vast majority of newly infected individuals progress into a latent stage where *M. tuberculosis* remains dormant in a non-replicative stage, and as such, non-infectious. Breaking out from the latent stage into active pulmonary disease is a process mediated by several risk factors, and one that in most cases never realizes. At the population level it is accepted as a broad estimation that 10% of latent infections will break into active disease during a life-span (2). On the other hand, we also know that \~10% of new TB transmission events will fast-track into active TB disease (3). Active TB cases, as we discussed above, are symptomatic and infectious. Intermediate stages of subclinical active TB are currently recognized, but for simplicity we will accept here that symptomatic disease equals infectiousness.

The average duration of the infectious period is \~3 years. Cohort studies have established that after this period roughly 50% of active TB cases die from the disease, while the other 50% recover spontaneously (4). TB does not confer complete immunity after recovery, however partial immunity has been observed in previously infected and might protect from reinfection (HR \~0.5) (5). Finally, individuals who had recovered from active TB are not only susceptible for re-infection, but they can relapse into active disease, usually at a yearly rate of around 0.5% (6).

**1) Following the statement above, can you complete the table of parameters? Draw a flow chart if necessary.**\
*(Note: All parameters with prefix T. are in a scale of time (years). Use this parameters to calculate the missing rates as necessary).*

First, let's load some libraries we need.

```{r echo=TRUE}
#Load packages
library(deSolve)
library(gridExtra)
library(ggplot2)
library(reshape)
library(R6)

```


```{r }

# Model Parameters
T.lfx    <- 72              # Life expectancy
T.dur    <- 3               # Duration of infectious period
mu       <- 1/T.lfx         # Background mortality rate
mu_tb    <- 0.5/T.dur     # TB mortality rate
imm      <- 0.5             # Infectiousness decline (partial immunity)
r_slow   <- 0.5             # rate from latent fast to latent slow
relapse  <- 0.005           # Relapse rate
r_incar  <- 200/1e5         # Incarceration rate
r_reincar<- 1000/1e5        # Re-Incarceration rate
r_release<- 0.5             # release rate (1/mean term time) 
RRc      <- 2.1               # Ratio of increase in contact in prison relative to community
RRbreak_p<- 9               # relative increase in TB reactivation in prison
RRbreak_x<- 3               # relative increase in TB reactivation in prison

#Community params
beta     <- 5               # Transmission rate per capita
break_fast <- 0.01          # Transition rate from latent to active disease
break_slow <- 0.008         # Transition rate from latent to active disease
selfcure <- 0.5/T.dur     # Rate of spontaneous cure

#Prison params
beta_p     <- beta*RRc         # Transmission rate per capita
break_fast_p <- break_fast*RRbreak_p # Transition rate from latent to active disease
break_slow_p <- break_slow # Transition rate from latent to active disease
selfcure_p <- 0.5/T.dur      # Rate of spontaneous cure

#Ex Prison params
break_fast_x <- break_fast*RRbreak_x # Transition rate from latent to active disease
break_slow_x <- break_slow # Transition rate from latent to active disease
selfcure_p <- 0.5/T.dur      # Rate of spontaneous cure


```

```{r }
# Load the R functions for TB transmission in prisons and for analysisng interventions
source('R/TB_prison.R')
source('R/useful_functions.R')
```

```{r }
t.intervention <- 400      # years of simulation
t.scale        <- 3        # Scaling up time of interventions
times <- seq(0, t.intervention, by = 1)          # time scale
```


```{r }

# Create
params <- c(
    break_fast = break_fast,
    break_fast_x = break_fast_x,
    break_fast_p = break_fast_p,
    break_slow = break_slow,
    break_slow_x = break_slow_x,
    break_slow_p = break_slow_p,
    selfcure = selfcure, 
    selfcure_x = selfcure_x, 
    selfcure_p = selfcure_p, 
    mu = mu, 
    mu_tb = mu_tb,
    beta = beta,
    beta_p = beta_p, 
    r_slow=r_slow,
    imm = imm, 
    relapse = relapse,
    r_incar=r_incar,
    r_reincar=r_reincar,
    r_release=r_release) 

# Initial Conditions
N  <- 1                    # Total population equal to unity 
I0 <- 1e-6                 # TB seed at time 0
Np  <- 0.005               # Total population equal to unity 
I0p <- 0.005*1e-6*10        # TB seed at time 0
Nx  <- 0.005*0.5           # Total population equal to unity 
I0x <- 0.005*0.5*1e-6*5        # TB seed at time 0


xstart <- data.frame(
                     U = N-I0-Np-Nx,
                     Lf = 0,
                     Ls = 0,
                     I = I0,  
                     R = 0,
                     Ux = Nx-I0x,
                     Lfx = 0,
                     Lsx = 0,
                     Ix = I0x,  
                     Rx = 0,
                     Up = Np-I0p,
                     Lfp = 0,
                     Lsp = 0,
                     Ip = I0p,  
                     Rp = 0,
                     Incidence = 0, 
                     Irecent = 0, 
                     Iremote = 0,
                     Incidence_x = 0, 
                     Irecent_x = 0, 
                     Iremote_x = 0,
                     Incidence_p = 0, 
                     Irecent_p = 0, 
                     Iremote_p = 0)     
```



```{r }
out  <- get_intervention(xstart, params, NA, times, NA, NA, TB_prison,
"Initial", NA) 

# plot
out$inc_all
               
```


```{r echo=TRUE, warning=FALSE}
# Function handles

fx_basic<-TB.Basic
fx_scale<-function(t, state, parameters) scale_up(t, state, parameters,t.interv,params,fx_basic)
```

Now, lets create a baseline incidence to use as counterfactual in our exploration of TB elimination.

*Copy and Paste the code below*

```{r echo=TRUE, warning=FALSE}
## Simulation 0
# Project a baseline
int_name   <- "Baseline"

#Initial conditions (starting in 2019)
sfin       <- tail(out0$out,1)                             
params_base<- params
times_new  <- seq(t.intervention, t.intervention+25 , by=1)
t.interv   <- c(times_new[2], times_new[2]+t.scale)

#Run model
data0<-get_intervention(sfin, params_base, params_base,times_new,
                             NA, fx_scale,fx_basic, "Baseline", NA) 
#Plot
grid.arrange(data0$lines ,data0$pie )
 


```

## *TB treatment*

We will start our exploration of interventions by simulating the roll-out of a successful TB treatment campaign. For simplicity, we want to do so without altering our current model structure. The simplest way is to alter the rate of spontaneous cure.

In the pathway to successful (curative) TB treatment, there are a number of events an individual should go through.

a)  Careseeking rate (cs): We can define careseking as the time it takes for a symptomatic individual to seek for medical care
b)  Probability of diagnosis (pDx): the probability that once an individual has sought care, a diagnostic test will be performed
c)  Probability of treatment (pTx): probability that once an individual has been diagnosed, treatment will be prescribed
d)  Treatment duration (T.rtx): Total duration of the treatment course (standard TB treatment is 6 months)

For the sake of simplicity we will not include other factors that might also play a role in treatment success.

**5) Can you list at least two other factors related to TB treatment success?**

<!-- --- -->

<!-- ####*Answer* -->

<!-- 5)  -->

<!-- * Treatment completion rates: TB treatment is long and potentially toxic. Low adherence is a big problem for TB programmes -->

<!-- * Drug sensitivity of TB strain: even with full completion, resistance to components of TB treatment regimen will invariably lead to treatment failure     -->

<!-- * Extra: Concomitant TB and HIV treatment: even with full completion and a drug-sensitive TB  strain, interactions with antiretrovirals might reduce the bioavailability of TB drugs.        -->

<!-- --- -->

**6) In the code below, assign values to the variables that reflect the components of TB treatment. Consider an average careseeking delay of 1 year, a probability of diagnosis of 95%, and treatment initiation of 95%.**

**7) Write the code for the term Tx. (Note that Tx is a rate that will add to the existing self cure rate in your model). Run the code to see your results.**

```{r eval=FALSE, warning=FALSE}
## Simulation 1
# An Intervention simulating introduction of treatment
int_name   <- "Treatment"

#Update parameter results data frame to append new results
params_1 <-params_base
data_stub<-data0$data

#:::: COMPLETE THE MISSING VALUES AND WRITE A TERM FOR RATE Tx

# Change parameters for intervention
T.cs <-       # Time delay (yrs) between developing symptoms and seeking for care
pDx  <-       # Probability of being diagnosed once sought care
pTx  <-       # probability of recieving correct Tx if diagnosed
T.rTx<-   0.5 # 6 months treatment duration
Tx <-

#::::
  
params_1["selfcure"]<-selfcure + Tx

data1<-get_intervention(sfin, params_1, params_base,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data1$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data1$pie )
```

You should be looking at a plot similar to the one below.

**8) What can you say about the proportion *Recent vs. Remote* ?**

```{r echo=FALSE, warning=FALSE}
## Simulation 1
# An Intervention simulating introduction of treatment
int_name   <- "Treatment"

#Update parameter results data frame to append new results
params_1 <-params_base
data_stub<-data0$data


# Change parameters for intervention
T.cs <- 1     # Time delay (yrs) between developing symptoms and seeking for care
pDx  <- 0.95   # Probability of being diagnosed once sought care
pTx  <- 0.95   # probability of receiving correct Tx if diagnosed
T.rTx<- 0.5 # 6 months treatment duration

Tx <-pDx*pTx*(1/(T.cs+T.rTx))

params_1["selfcure"]<-selfcure + Tx

data1<-get_intervention(sfin, params_1, params_base,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data1$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data1$pie )
```

<!-- --- -->

<!-- ####*Answer* -->

<!-- 8) The fraction of incident cases from a remote source has increased with the intervention. This reflects the fact that successful TB treatment is removing individuals from the infectious pool at a higher rate than they are generated.     -->

<!-- ---    -->

As you could see from the plot above, we have included dashed lines for WHO's End TB goal (black) and a TB elimination threshold (red). This is so we can check how our model performs against these reference goals. WHO has define the End TB goal as a 95% reduction in TB incidence by 2035. Elimination is defined as a threshold \<1 case per million population. The latter should be interpreted as the theoretical limit for eradication, while the End TB is meant to be an achievable goal, which, if achieved, should leave countries at the door of elimination.

## *Increase the demand for TB services*

Now, let's combine the treatment campaign with an intervention that make TB services widely available in the community. This intervention should increase the current yield of TB prevalent cases screened by 75%.

**9) What parameters in our previous simulation would you alter to simulate such intervention? Modify the code below and run.**

*(Note: remember that we want to combine interventions, i.e., the previous intervention should be introduced here too)*

```{r eval=FALSE, warning=FALSE}
## Simulation 2
# An Intervention simulating demand generation
int_name   <- "Demand generation"

#Update parameter results data frame to append new results
params_2 <-params_1
data_stub<-data1$data

#::::::: COMPLETE THE MISSING VALUES AND WRITE A NEW TERM FOR Tx

# Change parameters for intervention
T.cs <-      # Time delay (yrs) between developing symptoms and seeking for care
pDx  <-      # Probability of being diagnosed once sought care
pTx  <-      # probability of receiving correct Tx if diagnosed
T.rTx<-      # 6 months treatment duration

Tx <-

#:::::::::::
  
  
params_2["selfcure"]<-selfcure + Tx

data2<-get_intervention(sfin, params_2, params_base,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data2$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data2$pie )
```

Our plot should look like this..

```{r echo=FALSE,, warning=FALSE}
## Simulation 2
# An Intervention simulating demand generation
int_name   <- "Demand generation"

#Update parameter results data frame to append new results
params_2 <-params_1
data_stub<-data1$data

# Change parameters for intervention
T.cs <-  1*(1-0.75)    # Time delay (yrs) between developing symptoms and seeking for care
pDx  <-  0.95     # Probability of being diagnosed once sought care
pTx  <-  0.95     # probability of receiving correct Tx if diagnosed
T.rTx<-  0.5     # 6 months treatment duration

Tx <-pDx*pTx*(1/(T.cs+T.rTx))

params_2["selfcure"]<-selfcure + Tx

data2<-get_intervention(sfin, params_2, params_BASE,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data2$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data2$pie )
```

We can see from our remote/recent pie chart that even after a strong combined campaign to reduce TB burden, some transmission remains.

## *TB transmission control*

Let's imagine a hypothetical scenario where transmission is completely shut down. This is practically unfeasible, but for the sake of testing our elimination case let's imagine we can stop transmission with a fully curative intervention in every single prevalent case.

**10) Run the code below.**

*Copy and Paste the code below*

```{r echo=TRUE, warning=FALSE}
## Simulation 3
int_name<-"Transmission stop"

#Update parameter results data frame to append new results
params_3 <-params_2
data_stub<-data2$data

# An Intervention simulating transmission reduction

params_3["beta"]<-0


data3<-get_intervention(sfin, params_3, params_base,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data3$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data3$pie )

```

A fully curative intervention appears insufficient to drive the TB epidemic to elimination.

**11) If transmission has stopped (as seen in the pie chart), where is the remaining incidence coming from?**

**12) Where should we intervene to finally drive the TB epidemic down?**

<!-- --- -->

<!-- ####*Answer* -->

<!-- 11) Turning beta to zero will stop ongoing transmission but relapse is still happening among previously infected and latent infections are still breaking into active disease.  -->

<!-- 12) Relapse could only be prevented by introducing a therapy that is 100% sterilizing which at the moment is not available. But the latent pool of infected can be addressed with preventive treatment, which is currently available.   -->

<!-- --- -->

## *LTBI treatment*

Let's introduce a prevention campaign that treats 100% latent TB infection (LTBI) and reduces progression into active TB in 99%.

**13) Run the code below**

*(Note that the black dashed line below represent the End TB goal and the read dashed line is the elimination threshold)*

*Copy and Paste the code below*

```{r echo=TRUE, warning=FALSE}
## Simulation 4
# An Intervention simulating LTBI treatment
int_name<- "Prevention"

#Update parameter results data frame to append new results
params_4 <-params_3
data_stub<-data3$data

params_4["break_in"]<-0.01*(1/T.lfx)

data4<-get_intervention(sfin, params_4, params_base,times_new,
                       t.interv, fx_scale,fx_basic, int_name, data_stub) 

p1<-data4$lines +
  # EndTb
  geom_hline(yintercept=Inc.country*0.1, linetype="dashed", color = "black", size=1)+
  # Elimination
  geom_hline(yintercept=0.1/1e5, linetype="dashed", color = "red", size=1)


grid.arrange(p1 ,data4$pie )

```

This final intervention scenario reach the goal for End TB just before 2035 but is still far from elimination.

We have coded and run a simple TB model for exploring the case of TB elimination and WHO's End TB goal. Even though we did not formally assessed this subject, our simple exercise suggests that a combination of curative and preventive measures will be required to achieve at least End TB goals in a high burden setting.

Finally, take a few minutes to think how the following added complexities to the model could affect our estimations:

a)  Age structure

b)  Risk groups (e.g. HIV, Slum dwellers, diabetes, malnourished)

c)  MDR-TB

<!-- --- -->

<!-- ####*Answer* -->

<!-- a) Age is an important factor of TB dynamics: Children are more likely to develop extra-pulmonary TB (non-infectious) which overall makes this group (0 to 5 years) less likely to transmit the infection. Also, as the slow progression to latent TB can take periods as long as a life time, it is common to observe that reduction of burden will first impact younger ages while the older groups accumulate the larger number of latent progressors. Finally, age mixing patterns are an important factor when establishing the overall size of the epidemic.  -->

<!-- b) Transmission and progression to active TB is known to be strongly related to risk factors that drive epidemics in different settings. Existence of risk groups means that control strategies can be drastically reduced in scale if targeted to those risk groups fuelling the epidemic. -->

<!-- c) Since MDR-TB is the resulting effect of a wide access to TB drug regimens. As such, incomplete coverage and treatment completion will effectively lead epidemics to transition into higher rates of MDR-TB, making the case of elimination even harder.    -->

<!-- --- -->

------------------------------------------------------------------------

## References

(1) [World Health Organization . Global Tuberculosis Control: WHO Report 2011. Geneva: WHO; 2011.](https://apps.who.int/iris/handle/10665/44728)

(2) [Horsburgh CR, O'Donnell M, Chamblee S, Moreland JL, Johnson J, Marsh BJ, et al. Revisiting rates of reactivation tuberculosis: a population-based approach. Am J Respir Crit Care Med. 2010;182(3):420-5.](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2921602/)

(3) [Vynnycky E, Fine PE. The natural history of tuberculosis: the implications of age-dependent risks of disease and the role of reinfection. Epidemiol Infect. 1997;119(2):183-201.](https://www.ncbi.nlm.nih.gov/pubmed/9363017)

(4) [Tiemersma EW, van der Werf MJ, Borgdorff MW, Williams BG, Nagelkerke NJ. Natural history of tuberculosis: duration and fatality of untreated pulmonary tuberculosis in HIV negative patients: a systematic review. PLoS One. 2011;6(4):e17601.](https://www.ncbi.nlm.nih.gov/pubmed/21483732)

(5) [Andrews, J.R., et al., Risk of progression to active tuberculosis following reinfection with Mycobacterium tuberculosis. Clin Infect Dis, 2012. 54(6): p. 784-91.](https://www.ncbi.nlm.nih.gov/pubmed/22267721)

(6) [Thomas A, Gopi PG, Santha T, Chandrasekaran V, Subramani R, Selvakumar N, et al. Predictors of relapse among pulmonary tuberculosis patients treated in a DOTS programme in South India. Int J Tuberc Lung Dis. 2005;9(5):556-61.](https://www.ncbi.nlm.nih.gov/pubmed/15875929)

------------------------------------------------------------------------

#About this document

## Contributors

-   [Juan F. Vesga](https://www.imperial.ac.uk/people/j.vesga10)

The source file is hosted on [github](https://github.com/juanvesga/TBpractical/blob/master/tb_practical.Rmd).
